---
title: "Midterm"
author: "Mohammad Shehab"
date: "4/1/2020"
output:
  html_document: default
  pdf_document: default
---

# Introduction


On April 15, 1912, the largest passenger liner ever made collided with an iceberg during her maiden voyage. When the Titanic sank it killed 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships. One of the reasons that the shipwreck resulted in such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others.

The titanic.csv file contains data for 1310 of the real Titanic passengers. Each row represents one person. The columns describe different attributes about the person including whether they survived (S), their age (A), their passenger-class (C), their sex (G) and the fare they paid (X).

We are going to perform a Exploratory Data Analysis to try to understand the survivors pattern.How was the classes distributes and does it means if you are on higher class you had more chances of survival ?

--- 

# EDA Goal

We are going to perform an exploratory data analysis to our titanic dataset, we will ask some questions about our data set to discover the variables and values, then we will visualize the distributions perform some transformation and then the model to come up with conclusion and analysis. 

---

# Dependencies 

In order to execute our program we have to have the following dependencies. I will not be loading the titanic dataset from a csv file however i have installed a package to do this for me.

```{r}
#install.packages("tidyverse")
#install.packages("rpart.plot") 
#install.packages("caret")
#install.packages("e1071")
```

```{r}
library(tidyverse)
library(rpart.plot) 
library(caret)
```

#Loading the Titanic data

```{r}
#clean the workspace for every run
rm(list = ls())
titanic <- read.csv("titanic.csv", na.strings="")
colnames(titanic)
```

# Framing Questions 

- 1 - What is the age distrbution on the ship ? 
- 2 - How the age distrubuted between sex and age ?
- 3 - How did the age and sex affected the survival rate ?
- 5 - Does being in a specific class affects your survival rate? Can this apply for ages alike ?
- 4 - How was the fare distributed?

# Visualisation of Distribution

- we will examin the distribution of the passenger's sex.

```{r}
# if we want to filter `na` we can simply say
# titanic<-titanic %>%
#  filter(!is.na(Sex))
# however, we are not cleaning the data.

ggplot(data = titanic) +
    geom_bar(mapping = aes(x=Sex, fill=Sex), col = "black", alpha = 0.8) + 
    labs(x = "Sex", y = "Count", title = "Sex Distribution") +
    theme_classic()
```

Since sex is a categorical variable we use a bar chart. The bar chart shows that there is 1 `NA` sex which can indicate that the gender is missing for the individual, the bar chart displays that the number of males is greater than females.

We can display the following by using the `count`. The count function generates an error when using `NA` with factors so we remove the `NA`

```{r}
titanic<-titanic %>%
 filter(!is.na(Sex))
```

```{r}
titanic %>%
  count(Sex)
```

To examin the distribution of the `Age` we will use a histogram chart.

```{r}
ggplot(data = titanic) +
      geom_histogram(mapping = aes(x = Age), binwidth = 0.5) + 
      labs(x = "Age", y = "Count", title = "Age Distribution") +
        theme_classic()
```

The `age` distribution is plotted above in a histrogram which shows a `positive skewed` distribution, this can give an indication on the median if we plot this on a box and whiskers.

```{r}
 ggplot(data = titanic, mapping = aes(x = Age), horizontal = T, main= "Age Distribution of Passengers") +
   labs(x = "Age", title = "Age Distribution along") +
      geom_boxplot()
```

This confirms the following: The lower quartile of the box is 20 which means 25% where under the age of 19, and 75% where older than the age of almost 25. Although we had some outliers which were above the age of 60+. The mean will not give an accurate indication as it has been affected by the `outliers` and the `NA's`
if we calculate the median the below results confirms that the avarage age 28 where the majority `75%` of the passangers who were on the titanic. 
```{r}
summary(titanic$Age)
```

the following observation if we calculate the number of people with age respectively and display them in a tabular format. 

```{r}
titanic %>%
      count(cut_width(Age, 0.5))
```


It will indicate that the 42 passengers are between the age of `[23.8 and 24.2]`. if we to look at the lower bound of the box we can have something that looks like this

```{r}
young<- titanic %>%
    filter(Age<26.8)

## then we can plot data
ggplot(data = young) +
      geom_histogram(mapping = aes(x = Age)) + 
      labs(x = "Age", y = "Count", title = "Passengers Under 26.8 years of age") +
        theme_classic()
```


we notice how quickly the observation have changed, now it's showing a `negatively skewed distribution` towards the right.If we use a histrogram and correlate with the Age/Sex it might give us a better indication.Hence, we can observe that the Males younger on the ship are more in numbers with comparison the females.

```{r}
ggplot(data = young, mapping = aes(x = Age, color = Sex)) +
      geom_freqpoly(binwidth = 1)
        labs(x = "Age", y = "Count", title = "Passengers Under 26.8 years of age") +
          theme_classic()
```


We now observe how the survival rate was disributed overall between ages.
it almost makes sense as Females had more chances in surviving than males based on the age distribution along with older females had more chances of survival females who are below the age of 40 had a 75% chance for survival.

```{r}
ggplot(titanic,aes(Survived, Age, fill = Sex))+
  geom_boxplot()
```


We now observe the class with age.

```{r}
ggplot(data = titanic , aes(x = as.factor(Passenger.Class), 
                            y = Age, colour = Survived)) +
geom_boxplot()
```

We observe that people in the first class had a higher chances of survival however only young people had a higher chances people who were in the first class and above the age of 38 had 50% chances of surviving on the other hand people of where above 60 of age had less than 25% chances or surviving.


- How was the fare distributed?

```{r}
ggplot(data = titanic) + 
  geom_point(mapping = aes(x = Passenger.Class, y = Passenger.Fare), alpha = 1 / 100)
```

Very normal to have fares high based on the class; However we observe that some people paid very high amount for the same class.

```{r}
ggplot(data = titanic , aes(x = Passenger.Class,y = Passenger.Fare, colour = Sex)) +
  geom_boxplot()
```
`WOW` that was not expected, females with same class as men paid higher amount for the fare!! some outliers paid 500 for the fare and he is with the same class as the others.

## Refining and Generating More Questions

The following exploration leads us to generate more questions.

-  Anything is unusual about the age?
-  What is the survival rate for elderly?
-  What was the survival ratio between the sex in the elderlies ?

we use the `coord_cartesian` to zoom in and see what in unsual about some points. In the following nothing is unusal about the points except for 1 passenger who was 80 years of age.


```{r}
ggplot(titanic) + 
  geom_histogram(mapping = aes(x = Age), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 2))
   
unusual <- titanic %>% 
  filter(Age < 1 | Age > 60) %>% 
  select(Age) %>%
  arrange(Age)
unusual
```



Now we will try to observe the survival rate between the different ages and sex. but before that we will `mutate`
our data and create an `AgeGroup` column which is a calculation of ages.

```{r}
age_groups <- titanic %>%
  mutate(AgeGroup = 
           ifelse(Age < 12, 'Child', 
                  ifelse ( Age > 12 & Age < 60, 'Adult', 
                           ifelse( Age > 59, 'Elderly', NA)))) %>%
    select(Survived, Age, AgeGroup, Sex) %>%
  arrange(Age)

## plot it
ggplot(data = age_groups, mapping = aes(x = Age, y = AgeGroup, color=Survived)) + 
  geom_point()
```

This is very new information for me I would have thought that Elderly are to be saved first (Children, women, and elderly) this is clearly not the case here. we can plot this on a box and whiskers and see how the distribution of elderly and correlate it with the class

we can prove this by running the following which shows 28 elderly did not survive.

```{r}
all_elderly <- age_groups %>%
  filter(Age > 59)  %>%
    group_by(Survived) %>%
    count(AgeGroup)
```

```{r}
age_groups %>%
  ggplot(mapping = aes(Age)) + 
    geom_freqpoly(mapping = aes(colour = Sex), binwidth = 1/4)
```

Clearly the diagram shows that more aged males were available on the ship than women. But what would be the survival rate between those?

```{r}
age_groups <- titanic %>%
  filter(Age > 59) %>%
  select(Survived, Age, Sex, Passenger.Class) %>%
  arrange(Age)

ggplot(data = age_groups , aes(x = Passenger.Class, y = Age, colour = Survived)) +
  geom_boxplot()
```

The box plot indicates that the old aged in the first class who were equal or above 60 survived. On the other hand almost 75% under the age of 66 did not survive in the first class. In the second and third class however no survival rate was available between the elderlies. So if you were an old aged man on the titanic you had almost 0 chance of survival if you were in the 2nd or 3rd class.

## Conclusion:

This is clearly just the begining of not even scratching the surface of the `titanic` dataset I had no idea that the elderly would have a very chance survival rate and I would have always expected that children women and elderly clearly that was not the case on the titanic. Fare distribution was again was a WOW factor for me males paid cheaper amount for the same class than females!








