---
title: "Midterm"
author: "Mohammad Shehab"
date: "4/1/2020"
output: html_document
---

# Introduction


On April 15, 1912, the largest passenger liner ever made collided with an iceberg during her maiden voyage. When the Titanic sank it killed 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships. One of the reasons that the shipwreck resulted in such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others.

The titanic.csv file contains data for 1310 of the real Titanic passengers. Each row represents one person. The columns describe different attributes about the person including whether they survived (S), their age (A), their passenger-class (C), their sex (G) and the fare they paid (X).

We are going to perform a Exploratory Data Analysis to try to understand the survivors pattern.How was the classes distributes and does it means if you are on higher class you had more chances of survival ?


--- 

# Dependencies

In order to execute our program we have to have the following dependencies. I will not be loading the titanic dataset from a csv file however i have installed a package to do this for me.

```{r}
#install.packages("tidyverse")
#install.packages("rpart.plot") 
#install.packages("caret")
#install.packages("e1071")
```

```{r}
library(tidyverse)
library(rpart.plot) 
library(caret)
```

#Load and Describing our data and a quick summary

```{r}
#clean
rm(list = ls())
train <- read.csv("titanic.csv", na.strings="")
str(train)
```

# Data Preparation and Cleaning

I'm trying to add more attributes to the model in such i dont want to use strings when you have true/false or male female or survived or not it's always better using 1 character or boolean vs using a string.

I will also try to remove any nullabale values as it makes it much more easier in the data processing.

```{r}
# survived
train$SurvivedFac <- ifelse(train$Survived=="Yes",1,0)
train$SurvivedNum <- as.numeric(train$SurvivedFac) 

# class
train$PclassFac <- ifelse(train$Passenger.Class=="First", "1st Class", ifelse(train$Passenger.Class=="Second", "2nd Class", "3rd Class"))
train$PclassFac <- factor(train$PclassFac)
train$PclassNum <- ifelse(train$Passenger.Class=="First", 1, ifelse(train$Passenger.Class=="Second", 2, 3))


#adding the length of the name as it might mean something
train$NameLength <- vector("numeric", nrow(train))
for (i in 1:nrow(train)) {
    train$NameLength[i] <- nchar(as.character(train$Name)[i])
}

train$SexFac <- ifelse(train$Sex=="Male", 1,0)

summary(train)
```

#Data visualization

1 - I'm trying to correlate between the length of the name (which might be the nationality) and the fare amount in pound. it might give an idea about the distribution of the nationalities. Which means if you had a longer name you might afford a pricy ticket? probably and might be irrelevant.

```{r}
# see whether NameLength is useful
plot(train$NameLength, train$Passenger.Fare, 
    pch=19, col=rgb(0,0,1,alpha=0.2),
    xlab="Name Length (chars)", 
    ylab="Fare (Pounds)")
abline(lm(train$Passenger.Fare ~ train$NameLength), col="red")
```

- Let's take a look at the age distribution. And we might create a correlation between males and females.

```{r}
ggplot(data=train, aes(x=Age)) +
  geom_histogram(colour="black", fill="seagreen4",binwidth=10) +
  ggtitle("Age Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.minor = element_line(colour = "red", linetype = "dotted"),
        panel.background = element_rect(colour = "darkblue"))
```


 - So here we show how many `females` (survived or not) vs how many `males` (survived or not)

```{r}
# using the SurvivedNum as this is should be continuas
ggplot(data=train, aes(x=SurvivedNum, fill=Sex)) +
  geom_bar(width=0.7) +
  scale_fill_manual(values=c("hotpink", "deepskyblue"))+
  theme(panel.grid.minor=element_blank(),panel.grid.major=element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c("No", "Yes"))
```

 - Lets categories the passengers.

```{r}
 ggplot(train %>%
   filter(!is.na(Age))) + 
    geom_bar(aes(x = Passenger.Class), fill="#FF6666")
```

```{r}
ggplot(train %>%
   filter(!is.na(Age))) + 
  geom_bar(aes(x = Sex, fill = Sex))
```

Let’s create a plot to better understand the distribution of ages of passengers on the Titanic. A histogram is a common way to visualize the distribution of a continuous variable. Creating a histogram is a simple modification of our earlier examples where we created bar plots.

```{r}
 ggplot(train) +
   geom_histogram(aes(x = Age), bins = 35)
```

The histogram provides a quick visual representation of the frequency of different age groups. It appears that the the most common (modal) value of age is a little over 20 years old. We can explicitly calculate the mean and median age as follows:

```{r}
mean(train$Age, na.rm = TRUE)
```

```{r}
median(train$Age, na.rm = TRUE)
```


 - Take a look at our outlier. From this simple boxplot, it doesn’t look like there is a great difference in the age distribution of passengers who lived vs died.

```{r}
train %>%
   filter(!is.na(Age)) %>%
   ggplot() + 
   geom_boxplot(aes(Survived, Age))
```

- So Whats the probability for surviving males vs females? taking into consideration the age group.

```{r}
train %>%
  filter(!is.na(Age)) %>%
   ggplot(aes(x = Age, y = Survived, color = Sex)) + 
   geom_jitter(height = 0.05, alpha = 0.35) + 
   geom_smooth(method="glm", method.args = list(family="binomial"))  + 
   facet_wrap(~Sex) + 
   labs(x = "Age", y = "Probability of survival")
```

 - It seems that a lot of the middle aged passengers died while the younger age passengers of around age 1 or 2 are more likely to survive.

We can probably see it more clearly by looking at the survival comparison between Child and Adults using the AgeGroup feature we created from previous section.

```{r}
ggplot(train %>%
  filter(!is.na(Age))) +
  geom_freqpoly(mapping = aes(x = Age, color = Survived), binwidth = 1) +
  labs(x = "Age", y = "Number of Passengers", title = "Survival Comparison by Age") +
  theme_bw()
```

 - Lets take a look at the people who survived vs the people who didn't, do you remember the first point? I was pointing at the nationality as people who were poor coming from irland and different areas where sitting in class 3
they couldnt pay for the highest fare which included a life bought.

```{r}
## get the total count of the passengers

classCount <- train %>%
    group_by(PclassNum) %>%
    count(PclassNum) %>%
    select(PclassNum, Passengers = n)

## plot the pointss
ggplot(classCount, aes(x = PclassNum, y = Passengers)) +
    geom_bar(stat = "identity", aes(fill = PclassNum), col = "black", alpha = 0.8) +
    labs(x = "Passenger Class", y = "Number of Passengers", title = "Number of Passengers in each passenger class") +
    theme_classic()
```


--- 

# Final Project Stars Here

Lets take a look at the people in general.
Lets break this information a little. What does this means ? What if females where the ones who didnt survived the most ?
```{r}
#total Passangers
totalPassangers<- train %>%
                    count()

print(totalPassangers)

#All Genders
genders <- train %>%
  filter(!is.na(Sex)) %>%
    group_by(Sex) %>%
    count(Sex) %>%
    select(Sex, Passengers = n)

print(genders);     

# Ones who survived         
survived <- train %>%
  filter(!is.na(Sex)) %>%
    filter(Survived == "Yes")  %>%
    group_by(Sex) %>%
    count(Sex) %>%
    select(Sex, Passengers = n)

print(survived); 

## failed to survive
died <- train %>%
  filter(!is.na(Sex)) %>%
    filter(Survived == "No")  %>%
    group_by(Sex) %>%
    count(Sex) %>%
    select(Sex, Passengers = n)

print(died);  
```
The people who did'nt survived are the ones who where mostly males. So the entire ship was mostly male dominante. Does this means if we had more males the fitatlies would have been higher?

Looks like it as the out of 843 males only 161 survived this  is almost 64% of the entire passangers.

```{r}
ggplot(train %>%
  filter(!is.na(Sex)), aes(x = Survived, fill=Sex)) +
 geom_bar(position = position_dodge()) +
 geom_text(stat='count', 
           aes(label=stat(count)), 
           position = position_dodge(width=1), vjust=-0.5)+
 theme_classic()
```

Lets take a look how these passangers were distributed in terms of Class

```{r}
ggplot(train %>%
  filter(!is.na(Sex)), aes(x = Survived, fill=PclassFac)) +
 geom_bar(position = position_dodge()) +
 geom_text(stat='count', 
           aes(label=stat(count)), 
           position = position_dodge(width=1), 
           vjust=-0.5)+
 theme_classic()
```

Almost 77% of the men who died where in the lower class. It's very clear that the men who did not survived belonged to the lower classes of the ship which means they were sitting below the deck. This validates the story  that people who belonged to the upper class were preparing to leave the ship without telling the lower class.

# Age distribution 

```{r}
#data cleanup once and forall;

train<-train %>%
  filter(!is.na(Sex))
```

Lets take a look at the age distribution, most of the passengers where between the age group of 20 and 40. This is clearly a `positive` skew.
```{r}
ggplot(train, aes(x = Age)) +
 geom_density(fill='seagreen4')
```


Lets break it a little more lets try to see who are the people that survived by their age. 

```{r}
#filtering out  the people who did not have an age
train<- train %>%
        filter(!is.na(Age))

train$Discretized.age = cut(train$Age, c(0,10,20,30,40,50,60,70,80,100))
ggplot(train, aes(x = Discretized.age, fill=Survived)) +
  geom_bar(position = position_dodge()) +
  geom_text(stat='count', aes(label=stat(count)), position = position_dodge(width=1), vjust=-0.5)+
  theme_classic()
```

## Building The model.

I'm going to start by splitting my data into train data and testing data. Lets start preparing our train data.

My goal here is whenever i have new data i will give it to the model and it will try me if the new criteria will be in the survived or not class.

I'll start by creating a very simple `function` to prepare my data.

```{r}
#storing the train data in a tmp variable might need it later
#the fn trainMaker will create training data defaulted to 70% of my overall data
inTrain <- createDataPartition(y=train$Survived, p=0.70, list=FALSE)
train_ds <- train[inTrain,]
test_ds <- train[-inTrain,]
train_per <- (nrow(train_ds)/nrow(train))*100
test_per <- (nrow(test_ds)/nrow(train))*100
cat("Training dataset is : ", train_per, "% \n")
cat("Testing dataset is : ", test_per, "% \n")
```

## Building the decision tree

We are making a decision tree, the following is a classification done based on our training data.

```{r}
fit <- rpart(Survived ~ Sex + Age + PclassNum, data = train_ds, method = 'class')
rpart.plot(fit)
```

## Calculating Accuracy

How can we say that the following is correct? we begin by training the model. Here's how we do it. 

`Note` : we use our testing data for this to get the accuracy

```{r}
#train the model
model <- train(Survived ~ Sex + Age + PclassNum, method="rpart", data=train_ds)
#check the predicted accuracy
pred_train1 <- predict(model, newdata=train_ds)
print(confusionMatrix(pred_train1, train_ds$Survived))
table(pred_train1)
##Accuracy : 78%

## check the accuracy on the testing dataset  
pred_valid <- predict(model, newdata=test_ds) 
print(confusionMatrix(pred_valid, test_ds$Survived))
table(pred_valid)
##Accuracy : 77%
```


As per the above we have an accuracy of 77% of predecting a new value using the following model created above with a good confidence interval of 95%


```{r}
prepareData <-function(data){
  data$SurvivedFac <- ifelse(data$Survived=="Yes",1,0)
  data$SurvivedNum <- as.numeric(data$SurvivedFac) 
  
  # class
  data$PclassFac <- ifelse(data$Passenger.Class=="First", "1st Class", ifelse(data$Passenger.Class=="Second", "2nd Class", "3rd Class"))
  data$PclassFac <- factor(data$PclassFac)
  data$PclassNum <- ifelse(data$Passenger.Class=="First", 1, ifelse(data$Passenger.Class=="Second", 2, 3))
  
  #adding the length of the name as it might mean something
  data$NameLength <- vector("numeric", nrow(data))
  for (i in 1:nrow(data)) {
      data$NameLength[i] <- nchar(as.character(data$Name)[i])
  }
  data$SexFac <- ifelse(data$Sex=="Male", 1,0)
  data$Discretized.age = cut(data$Age, c(0,10,20,30,40,50,60,70,80,100))
  
  return (data)
}
```

### Testing data one last time

In this section I created some sample with few data. I created 8 entries with my name however i changed my age and the class. The model has predicted that out of these 8 entires only 3 will survive. Interesting...

```{r}
live <-prepareData(read.csv("./live.csv", na.strings=""));
predicted <- predict(model, newdata=live)
table<- table(predicted)
summary(table)
print(table)
```

# Clean Up

```{r}
#rm(list = ls())
```

---
# Conclusion

 
